An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-14 paper.tex
Minor edits, Makefile update.
Added 1,2,3 description of kD-tree algorithm.

From the set of points output by a single-model MCMC,
$\{\vtheta_i^{(j)} | j = 1, 2, \ldots \}$, all lying in the $M_i$
parameter space, we would like to construct a set of ``neighborhoods''
associated with each of the sample points.  The neighborhoods should
be non-overlapping and fill parameter space.  To draw a proposed jump
from an approximation to the model PDF, we select a point uniformly
from the $\vtheta^{(j)}_i$, find its associated neighborhood, and then
draw the proposed jump uniformly from the neighborhood.  Since the
MCMC points are distributed according to the posterior PDF for $M_i$,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  In fact, the proposed
jumps are drawn from a piecewise-constant (i.e.\ constant on each
neighborhood) interpolation of the PDF.

There are various techniques that could be used to construct the set
of neighborhoods associated with each point.
Ref.~\cite{Littenberg2009} decomposed parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each point was associated with the brick that
contained it.  Unfortunately, this process does not guarantee that
every brick contains at least one point, which is required for the
jump proposal to cover the entire model parameter space.  The solution
to this problem was to add a uniform distribution to the proposal PDF,
in the form of one ``fictitious'' point per brick.

>>>>

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
These neighborhoods should be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  In fact, the proposed
jumps are drawn from a piecewise-constant (i.e.\ constant on each
neighborhood) interpolation of the PDF.  There are various techniques
that could be used to construct the set of neighborhoods associated
with each point.

Ref.~\cite{Littenberg2009} decomposed parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each point was associated with the brick that
contained it.  Unfortunately, this process does not guarantee that
every brick contains at least one point, which is required for the
jump proposal to cover the entire model parameter space.  The solution
to this problem was to add a uniform distribution to the proposal PDF,
in the form of one ``fictitious'' point per brick.  The extra uniform
distribution from the fictituous points represents a significant cost,
since inter-model jumps proposed from the uniform distribution are
unlikely to be accepted.  Additionally, the bricks in this algorithm
do not adapt to the local structure of the PDF.  One must either use
small bricks to capture the local structure of the PDF, placing many
bricks in regions without MCMC samples, or use large bricks, missing
the local strucutre of the PDF in exchange for fewer empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.  A
larger difficulty with the Voronoi approach is that the computational
cost for computing the Voronoi regions for $N$ points increases
rapidly with dimension, $d$, as $N^X$ \will{Find X in the literature;
  add references.  I think it's $X = d/2$}.

We propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree
\will{References.}.  The decomposition is more adaptive than the cells
of Ref.~\cite{Littenberg2009}, and more efficient in high dimension
than the Voronoi decomposition.

A kD-tree is a binary, space-partitioning tree.  To partition a set of
points into a kD-tree, begin by placing them in a rectangular box that
contains all of parameter space.  Then proceed recursively:
\begin{enumerate}
\item If the given box contains exactly one point, stop; this is a
  leaf of the tree.  Otherwise:
\item Choose a dimension along which to divide the points.  Divide the
  points in half along this dimension (or nearly in half, if the
  number of points is odd), forming two sub-boxes.  The ``left''
  sub-box contains the half (or nearly half) of the points that have
  small coordinates along the chosen dimension; the ``right'' sub-box
  contains the half (or nearly half) of the points that have large
  coordinates along the chosen dimension.
\item Return to step 1 with each of the sub-boxes.
\end{enumerate}
